Assessing Bridge Condition in New York State Using Multiple Linear Regression

### 2024-12-04 paper/paper.qmd
Final update

---
title: "Assessing Bridge Condition in New York State Using Multiple Linear Regression"
subtitle: "Age, Municipality, and Ownership as Factors Influencing Bridge Condition Ratings "
author: 
  - Ariel Xing
thanks: "Code and data are available at: [https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis](https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis)."
date: today
date-format: long
abstract: "This study examines the factors influencing bridge conditions in New York State, focusing on the roles of age, municipality location, and ownership, using condition ratings provided by NYSDOT as a measure of bridge quality. Applying a multiple linear regression model, the analysis shows that for every additional year of age, a bridge's condition rating decreases by 0.02 points, reflecting the progressive impact of aging. On average, bridges located in towns and villages have condition ratings that are 0.17 and 0.19 points higher, respectively, than those in cities. Additionally, bridges owned by NYSDOT and other entities have condition ratings that are 0.09 and 0.16 points lower, respectively, than those owned by municipalities. These findings highlight the importance of targeted maintenance strategies that account for geographic and administrative factors to enhance infrastructure sustainability and safety.
"
format: pdf
toc: true 
number-sections: true
bibliography: references.bib
---

@@

This research highlights the importance of targeted maintenance strategies that address the specific needs of aging bridges across different geographic and administrative contexts. By identifying the primary factors affecting bridge conditions, the study supports more effective resource distribution and prioritization in infrastructure management. These results contribute to the long-term sustainability and safety of the state’s transportation network.

@@

This dataset was chosen over the updated 2020 version from New York State Open Data due to the change in how bridge condition data is recorded. In the 2016 dataset, the condition variable is a continuous variable measured on a seven-point scale (1 = "Deficient Condition" to 7 = "Good Condition"). However, in the 2020 dataset, NYSDOT replaced this scale with a binary Poor/OK variable, recorded as Poor01 (1 = Poor, 0 = OK). This change limits the analytical potential of the dataset, as the binary variable lacks the granularity required for linear regression that relies on continuous variables to detect subtle trends or incremental differences. By using the 2016 dataset, this study retains the detailed numerical condition ratings, enabling to analysis impact of factors such as age, location, and owners on bridge conditions.

@@

To conduct the analysis for this paper, we utilized the statistical programming language R [@citeR]. The workflow was supported by several packages, including tidyverse [@tidyverse] for data cleaning and manipulation, arrow [@arrow] for efficient data storage and retrieval, validate [@validate] for data testing and validation, knitr[@knitr] for creating tables and ggplot2 [@ggplot2] and broom[@broom] for creating visualizations.

@@

The model assumes that the relationship between predictors and the response variable is linear, and that the residuals are normally distributed with constant variance.

@@

The coefficient for Age is -0.02, indicating that for each additional year a bridge ages, its condition rating decreases by 0.02 points on average. This result reflects the natural deterioration of infrastructure over time, as older bridges are more likely to experience wear and tear. @fig-ageatinspection-condition presented in the Data section illustrates this trend, showing a clear negative relationship between bridge age and condition. This visual evidence aligns with the regression model, where the Age coefficient (-0.02) confirms this downward trend. Although the effect of age on condition appears small on an annual basis, the cumulative impact over several decades can significantly compromise a bridge's integrity, highlights the importance of regular inspections and maintenance programs to mitigate the impact of aging on bridge conditions.

@@

## Age Distribution Patterns {#sec-age-distribution}
The distribution of bridge ages at inspection shows distinct peaks approximately every 25 years, indicating a recurring cycle of bridge construction and major renovation. This pattern suggests that every 25 years, a significant number of bridges reach the point where extensive maintenance or replacement is necessary. The previous peak occurred roughly 25 years ago, meaning the next wave of large-scale repairs or new construction is approaching. Recognizing this pattern highlights the importance of early planning and budgeting to address the anticipated surge in infrastructure needs effectively, ensuring resources are ready to manage the simultaneous demands of aging bridges.

>>>>

---
title: "Assessing Bridge Condition in New York State Using Multiple Linear Regression"
subtitle: "Age, Municipality, and Ownership as Factors Influencing Bridge Condition Ratings "
author: 
  - Ariel Xing
thanks: "Code and data are available at: [https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis](https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis)."
date: today
date-format: long
abstract: "This study examines the factors influencing bridge conditions in New York State, focusing on the roles of age, municipality location, and ownership, using condition ratings provided by NYSDOT as a measure of bridge quality. Using a multiple linear regression model, the analysis indicates that for every additional year of age, a bridge's condition rating decreases by 0.02 points, reflecting the cumulative effects of aging. On average, bridges located in towns and villages have condition ratings that are 0.17 and 0.19 points higher, respectively, than those in cities. Additionally, bridges owned by NYSDOT and other entities have condition ratings that are 0.09 and 0.16 points lower, respectively, than those owned by municipalities. These findings highlight the importance of targeted maintenance strategies that account for geographic and administrative factors to enhance infrastructure sustainability and safety.
"
format: pdf
toc: true 
number-sections: true
bibliography: references.bib
---

@@

This research emphasizes the need for targeted maintenance strategies tailored to the specific requirements of aging bridges across diverse geographic and administrative contexts.By identifying the primary factors affecting bridge conditions, the study supports more effective resource distribution and prioritization in infrastructure management. These results contribute to the long-term sustainability and safety of the state’s transportation network.

@@

This dataset was chosen over the updated 2020 version from New York State Open Data due to the change in how bridge condition data is recorded. In the 2016 dataset, the condition variable is a continuous variable measured on a seven-point scale (1 = "Deficient Condition" to 7 = "Good Condition"). However, in the 2020 dataset, NYSDOT replaced this scale with a binary Poor/OK variable, recorded as Poor01 (1 = Poor, 0 = OK). This change limits the analytical potential of the dataset, as the binary variable lacks the granularity required for linear regression that relies on continuous variables to detect subtle trends or incremental differences. By using the 2016 dataset, this study retains detailed numerical condition ratings, enabling an analysis of the impact of factors such as age, location, and ownership on bridge conditions.

@@

To conduct the analysis for this paper, I utilized the statistical programming language R [@citeR]. The workflow was supported by several packages, including tidyverse [@tidyverse] for data cleaning and manipulation, arrow [@arrow] for efficient data storage and retrieval, validate [@validate] for data testing and validation, knitr[@knitr] for creating tables and ggplot2 [@ggplot2] and broom[@broom] for creating visualizations.

@@

The model assumes a linear relationship between predictors and the response variable, with residuals normally distributed and exhibiting constant variance.

@@

The coefficient for Age is -0.02, indicating that for each additional year a bridge ages, its condition rating decreases by 0.02 points on average. This result reflects the natural deterioration of infrastructure over time, as older bridges are more likely to experience wear and tear. @fig-ageatinspection-condition presented in the Data section illustrates this trend, showing a clear negative relationship between bridge age and condition. This visual evidence aligns with the regression model, where the Age coefficient (-0.02) confirms this downward trend. Although the effect of age on condition appears small on an annual basis, the cumulative impact over several decades can significantly compromise a bridge's integrity. This underscores the importance of regular inspections and maintenance programs to mitigate aging's effects on bridge conditions.

@@

## Age Distribution Patterns {#sec-age-distribution}
@fig-age-histogram shows the distribution of bridge ages at inspection has distinct peaks approximately every 25 years, indicating a recurring cycle of bridge construction and major renovation. This pattern suggests that every 25 years, a significant number of bridges reach the point where extensive maintenance or replacement is necessary. The previous peak occurred roughly 25 years ago, meaning the next wave of large-scale repairs or new construction is approaching. Recognizing this pattern emphasizes the need for proactive planning and budgeting to address the anticipated surge in infrastructure needs, ensuring resources are available to manage the simultaneous demands of aging bridges.
